---
title: "Gov 50 Final Project"
author: Ben Seklir
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---
```{r setup, include = FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
```


## Project thoughts

I am interested in exploring data related to the effect private equity ownership of nursing homes has on quality of care. I hypothesize that there might be a correlation between private equity ownership of nursing homes and lower quality of care, for the cost-cutting and restructuring tactics that these firms are associated with seem antithetical to the aims of an ideal nursing home. I find this topic to be particularly timely for two reasons. First, an increasing number of nursing homes are being acquired by private equity firms. Second, Covid-19 hit many nursing homes extremely hard, and as a result, these facilities fell under intense scrutiny.



## Proposal

How does the ownership type of nursing homes affect their quality? I hypothesize that nursing homes operated by for-profit organizations provide worse quality of care than those operated by non-profit organizations and the government. As I started to get at in my initial project thoughts above, I suspect that the reason for this comes down to the incentives of the different ownership types. I would expect that for-profit nursing homes would be looking to minimize expenses wherever possible (to maximize profit) and that the quality of care would take a hit as a result. My data, sourced from the Centers for Medicaid & Medicare Services (CMS), is a list of currently active nursing homes that includes both general information about the nursing home as well as its score across a number of quality metrics. The explanatory variable of interest is whether the nursing home is operated by a for-profit organization or a non-profit organization/the government. The variable can be found under the column titled "Ownership Type". My outcome variable a nursing home's overall quality, which can be found under the "Overall Rating" column. This variable is measured using CMS' "Five-Star Quality Rating System," which assigns a nursing home an aggregate score based on the its ratings for health inspections, quality measures, and staffing, with more stars indicating higher quality. If I observe that non-profit- and government-operated nursing homes have higher average overall rating scores than for-profit facilities, this would provide support for my hypothesis. If it turns out that for-profit nursing homes have higher or similar average overall rating scores than non-profit- and government-operated facilities, this would provide evidence against my hypothesis.



## Data visualization

```{r}
nursing_homes <- read_csv("NH_ProviderInfo_Oct2023.csv")

nursing_homes <- nursing_homes |>
  mutate(Umbrella = case_when(
    `Ownership Type` %in% c("For profit - Corporation", "For profit - Individual", "For profit - Limited Liability company", "For profit - Partnership") ~ "For-profit",
    `Ownership Type` %in% c("Non profit - Corporation", "Non profit - Other", "Non profit - Church related") ~ "Non-profit",
    `Ownership Type` %in% c("Government - Federal", "Government - State", "Government - County", "Government - City/county", "Government - City", "Government - Hospital district") ~ "Government",
    TRUE ~ "Other"
  ))

selected_columns <-  nursing_homes |> select(c('Provider Name','Umbrella','Overall Rating'))

rating_table <- selected_columns |>
  group_by(Umbrella) |>
  summarize(
    avg_overall_rating = mean(`Overall Rating`, na.rm = TRUE)
  )
print(rating_table)
```


